Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-04-03 Manuscript/manuscript.Rmd
results edits

We developed a computational framework, refered to as *TADRF*, aimed at determining the optimal set of data level characteristics necessary for TAD boundary prediction using Random Forest classification models. Our framework is resolution-specific and uses called TAD boundaries as ground truth to establish TAD-boundary and non-TAD boundary regions. Under our framework, we binned the linear genome at intervals specific to the Hi-C data resolution used to call TADs. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1 A). Under this construction, the number of non-TAD boundary regions are exceed the number of TAD boundary regions. To cope with the imbalance among the class of genomic bins, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. *Random over-sampling* (ROS) was defined as sampling with replacement from TAD boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

Various one-dimensional cell line-specific ChIP-seq tracks (genomic elements) were used to construct the feature space for the random forests including cis-regulatory elements, histone modifications, and transcription factor binding sites. Four different feature engineering procedures were developed for each genomic element (Figure 1 B). *Peak Signal* was defined as the [???include definition]. Counts of overlaps (*Overlap Counts*, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (*Overlap Percent*, OP) were defined as the proportion of a bin overlapped by an annotation region. *Distance* was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (Signal/OP), and spatial (distance) relationships between TAD boundaries and genomic annotations.

We applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material.

@@

@@

\newpage

@@

Our computational framework was used to identify the combination of data resolution, predictor type, and re-sampling technique necessary for optimal cell line specific TAD boundary prediction (Figure 2). The framework uses data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, using a particular predictor type *P = {Signal, OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}*. Once the model inputs were established, we trained a random forest classifier on chromosome-specific sections of the genome using 5-fold cross validation. The different combinations of inputs were evaluated by assessing the Matthews Correlation Coefficient (MCC) of each model. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 \times 21 (= 3360)$ models.

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 5-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data for unbiased model evaluation.](Figures/ensemble_framework.png)

>>>>

We developed a computational framework<!--, refered to as *TADRF*,--> aimed at determining the optimal set of data level characteristics necessary for TAD boundary prediction using random forest classification models. We then leveraged our optimally built random forest to precisely predict the genomic coordinates of TAD boundaries at base pair resolution.

Our framework is resolution-specific and uses arbitrary called TAD boundaries as ground truth to establish TAD-boundary and non-TAD boundary regions. To demonstrate, we applied our framework to publically available replicate Hi-C data sets from the GM12878 and K562 cell lines (Supplementary Table 1). Topologically associating domains (TADs) were subsequently called using the ARROWHEAD pipeline [@durand2016juicer] at five data resolutions: 5 kb 10 kb, 25 kb, 50 kb, and 100 kb. TAD boundary regions were defined as regions on the linear genome where a genomic bin contained a called TAD boundary ($Y = 1$), while non-TAD boundary regions were defined as genomic bins that did not contain a TAD boundary ($Y = 0$) (Figure 1A). Under this construction, the number of non-TAD boundary regions far exceed the number of TAD boundary regions. To cope with the imbalance among the classes of genomic bins, we evaluated three re-sampling techniques (in addition to no re-sampling) before predicting TAD boundaries. *Random over-sampling* (ROS) was defined as sampling with replacement from TAD boundary regions. *Random under-sampling* (RUS) was defined as sampling with replacement from non-TAD boundary regions. Lastly, we considered *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling [@chawla2002smote].

A total of 78 cell type specific functional genomic annotations were used to build the feature space of downstream random forest models (Supplementary Table 2). The annotations could be broken down into three sets including histone modifications (HM), cis-regulatory elements (CRE), and transcription factor binding sites (TFBS). Four different feature engineering procedures were developed for each genomic element (Figure 1 B). *Peak Signal* was defined as the [???include definition]. Counts of overlaps (*Overlap Counts*, OC) were defined as the total number of genomic annotation regions that overlapped within a given genomic bin. Percents of overlaps (*Overlap Percent*, OP) were defined as the proportion of a bin overlapped by an annotation region. *Distance* was defined as log2 transformed number of base pairs between between the centers of bins and regions defining genomic annotations. A log2-transformation of distance was used to account for the skewness present in the distributions of genomic distances (Supplementary Figure 1). These feature engineering procedures were performed to improve TAD boundary predictions by accounting for direct (OC), proportional (Signal/OP), and spatial (distance) relationships between TAD boundaries and genomic annotations.

@@

Thus, under our framework, we considered combinations of data for a particular cell line *L = {GM12878, K562}*, at a given resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, with a specific engineered feature space *P = {Signal, OC, OP, Distance}*, and re-sampling technique *B = {None, RUS, ROS, SMOTE}* (Figure 2). Model inputs were evaluated by assessing the balanced accuracy (BA) of each combination. Letting $|X|$ denote the cardinality of each set, we evaluated $|L| \times |R| \times |P| \times |B|$ different ensemble systems, totaling $2 \times 5 \times 4 \times 4 (= 160)$ models. The results presented are for the GM12878 cell line, with K562 results presented in the Supplementary material.

![**A diagram of the ensemble framework detailing the steps for combining different inputs used in the random forest model building pipeline.** Step 1 is to establish the rowise (genomic bins) and columnwise (genomic annotations) dimensions of the data matrix, **A**, by combining a resolution (5 kb, 10 kb, 25 kb, 50 kb, or 100 kb) and feature engineering procedure (overlap count (OC), overlap percent (OP), $log_{2}$ distance (Distance)). Step 2 is to split **A** into a training and testing set according to a 7:3 ratio. Step 3 is to apply a re-sampling technique to the training set (no re-sampling, random over-sampling, random under-sampling, or SMOTE re-sampling). Step 4 is to implement the random forest algorithm and perform 3-fold cross-validation to tune for the *mtry* parameter. Finally, step 5 is to validate the model on the separate test data and evaluate model performance using balanced accuracy (BA).](Figures/ensemble_framework.png)

@@

@@
